---
title: "Clean Up Gates"
format: html
---

# Background

The purpose of this [Quarto Markdown File](), is to clean up the .fcs files acquired as part of the [Baltimore Introductory Spectral Cytometry Course]() for just the cells that are of interest for signature characterization. This code originally derives from code featured as part of the [Cytometry in R]() course. 

# Getting Started

## Set Up

First, after confirming that they are installed, we need to attach any R packages that will be needed to our local computer environment. 

```{r}
library(dplyr)
library(stringr)
library(ggplot2)
library(flowWorkspace)
library(flowGate)
library(Luciernaga)
```

Specify where to find your fcs files, as well as where you want to store your pdf files once you create them. 

```{r}
Instrument <- "5L"
Day <- "1"

InstrumentFull <- paste0("Aurora", Instrument)
DayFull <- paste0("Day", Day)
StorageLocation <- file.path("data_2025", InstrumentFull, DayFull, "Raw", "Reference Group")

OutputLocation <- file.path("data_2025", InstrumentFull, DayFull, "Raw", "Reference Group_Cleaned")

if(!dir.exists(OutputLocation)){dir.create(OutputLocation)}
```

Checking our designated location, to make sure the .fcs files are actually there

```{r}
fcs_files <- list.files(StorageLocation, pattern=".fcs", full.names=TRUE)
fcs_files
```

We next need to distinguish between bead and cell controls

```{r}
Cells <- fcs_files[str_detect(fcs_files, "Cells")]
Beads <- fcs_files[str_detect(fcs_files, "Beads")]
```

## GatingSet

### Cells

In parallel, wrapping the required frameworks that are needed for gating. 

```{r}
CellsCS <- load_cytoset_from_fcs(Cells, truncate_max_range=FALSE, transformation=FALSE)
CellsGS <- GatingSet(CellsCS)
```

```{r}
SFC_Parameters <- colnames(CellsGS)
FluorophoresOnly <- SFC_Parameters[!stringr::str_detect(SFC_Parameters, "FSC|SSC|Time")]

Biexponential <-  flowjo_biexp_trans(channelRange=4096, maxValue=4194304,
  pos=5.62, neg=0, widthBasis=-1000)
MyBiexTransform <- transformerList(FluorophoresOnly, Biexponential)
```

```{r}
transform(CellsGS, MyBiexTransform)
```

### Beads

```{r}
#| eval: FALSE
BeadsCS <- load_cytoset_from_fcs(Beads, truncate_max_range=FALSE, transformation=FALSE)
BeadsGS <- GatingSet(Beads)
```

Transforming the data so everything is visible as we would in SpectroFlo

```{r}
#| eval: FALSE
SFC_Parameters <- colnames(BeadsGS)
FluorophoresOnly <- SFC_Parameters[!stringr::str_detect(SFC_Parameters, "FSC|SSC|Time")]

Biexponential <-  flowjo_biexp_trans(channelRange=4096, maxValue=4194304,
  pos=5.62, neg=0, widthBasis=-1000)
MyBiexTransform <- transformerList(FluorophoresOnly, Biexponential)
```

```{r}
#| eval: FALSE
transform(BeadsGS, MyBiexTransform)
```

## Gating

Assembling a Gating Template. 

### Cells

```{r}
CellGatingTable <- tibble::tribble(
  ~filterId,    ~dims,                          ~subset,      
  "scatter",    list("FSC-A", "SSC-A"),      "root",  
  "singlets",   list("FSC-A", "FSC-H"),      "scatter",  
)
```

```{r}
#| eval: FALSE
flowGate::gs_apply_gating_strategy(CellsGS, gating_strategy = CellGatingTable)
```

```{r}
#| eval: FALSE
gs_apply_gate_check(gs=CellsGS, gate="scatter")
```

```{r}
#| eval: FALSE
gs_apply_gate_check(gs=CellsGS, gate="singlets")
```

### Beads

```{r}
BeadGatingTable <- tibble::tribble(
  ~filterId,    ~dims,                          ~subset,      
  "scatter",    list("FSC-A", "SSC-A"),      "root",  
  "singlets",   list("FSC-A", "FSC-H"),      "scatter",  
)
```

```{r}
#| eval: FALSE
flowGate::gs_apply_gating_strategy(BeadsGS, gating_strategy = BeadGatingTable)
```

```{r}
#| eval: FALSE
gs_apply_gate_check(gs=BeadsGS, gate="scatter")
```

```{r}
#| eval: FALSE
gs_apply_gate_check(gs=BeadsGS, gate="singlets")
```



## Export

```{r}
RFolder <- here::here("R")
MyFunctions <- list.files(RFolder, full.names=TRUE)
purrr::walk(.x=MyFunctions, .f=source)
```

### Cells

```{r}
Data <- gs_pop_get_stats(CellsGS, "singlets")
Data
```

```{r}
summary(Data)
```

```{r}
AllDNs <- purrr::map(.x=CellsGS, subset="singlets", .f=Downsampling,
 DownsampleCount=178000, addon="", returnType="fcs",
 StorageLocation=OutputLocation)
```

