---
title: "Signature Matrix"
format: html
---

# Background

The purpose of this [Quarto Markdown File](), is to derive the signature matrix for the single-color control .fcs files acquired as part of the [Baltimore Introductory Spectral Cytometry Course]() for just the cells that are of interest for signature characterization. This code originally derives from code featured as part of the [Cytometry in R]() course. 

# Getting Started

## Set Up

First, after confirming that they are installed, we need to attach any R packages that will be needed to our local computer environment. 

```{r}
library(dplyr)
library(stringr)
library(ggplot2)
library(flowWorkspace)
library(flowGate)
library(Luciernaga)
```

Specify where to find your fcs files, as well as where you want to store your pdf files once you create them. 

```{r}
Instrument <- "5L"
Day <- "1"

NumberDetectors <- 64
UnstainedDetector <- "UV7"

InstrumentFull <- paste0("Aurora", Instrument)
DayFull <- paste0("Day", Day)
StorageLocation <- file.path("data_2025", InstrumentFull, DayFull, "Raw", "Reference Group")
OutputLocation <- file.path("data_2025", InstrumentFull, DayFull, "Raw", "outputs")

if (!dir.exists(OutputLocation)){dir.create(OutputLocation)}
```

Checking our designated location, to make sure the .fcs files are actually there

```{r}
fcs_files <- list.files(StorageLocation, pattern=".fcs", full.names=TRUE)
fcs_files
```

```{r}
TheCS <- load_cytoset_from_fcs(fcs_files, truncate_max_range=FALSE, transformation=FALSE)
TheGS <- GatingSet(TheCS)
```

```{r}
SFC_Parameters <- colnames(TheGS)
FluorophoresOnly <- SFC_Parameters[!stringr::str_detect(SFC_Parameters, "FSC|SSC|Time")]

Biexponential <-  flowjo_biexp_trans(channelRange=4096, maxValue=4194304,
  pos=5.62, neg=0, widthBasis=-1000)
MyBiexTransform <- transformerList(FluorophoresOnly, Biexponential)
```

```{r}
transform(TheGS, MyBiexTransform)
```

## Signature Matrix

```{r}
IntermediateMetadata <- pData(TheGS) |>
  mutate(
    name_clean  = str_remove(name, "^PD_AD_"),
    Antigen     = str_extract(name_clean, "^\\S+"),
    Fluorophore = str_match(name_clean, "^\\S+\\s(.*)\\s\\((?:Beads|Cells)\\)")[, 2],
    Type        = str_extract(name_clean, "(?<=\\()(?:Beads|Cells)(?=\\))"),
    Detector    = "",
    Negative    = "Unstained"
  )
     
IntermediateMetadata <- IntermediateMetadata |> 
  mutate(Antigen = case_when(
    Antigen == "1DeadCells" ~ "DeadCells",
    TRUE ~ Antigen
  )) 
  
IntermediateMetadata <- IntermediateMetadata |>
  mutate(Fluorophore = case_when(
    is.na(Fluorophore) ~ Antigen,
    TRUE ~ Fluorophore
  ))

IntermediateMetadata <- IntermediateMetadata |>
  mutate(Negative = case_when(
    Fluorophore == "7-AAD" ~ "DeadCells", # Consider flexibility
    TRUE ~ Negative
  )) 

IntermediateMetadata <- IntermediateMetadata |> mutate(
    Detector = purrr::map_chr(Fluorophore, \(f) DetectorGuess(x = f, NumberDetectors = NumberDetectors))
  )

IntermediateMetadata <- IntermediateMetadata |> select(-name_clean) |> relocate(name, Fluorophore, Antigen, Type, Detector, Negative)

IntermediateMetadata <- IntermediateMetadata |>
  mutate(Detector = case_when(
    str_detect(Fluorophore, "Unstained") ~ UnstainedDetector, # Consider flexibility
    TRUE ~ Detector
  )) 

IntermediateMetadata <- IntermediateMetadata |>
  mutate(Detector = case_when(
    str_detect(Fluorophore, "Dead") ~ UnstainedDetector, # Consider flexibility
    TRUE ~ Detector
  )) 

pData(TheGS) <- IntermediateMetadata
```

# Create our peak detectors gates 

```{r}
#| eval: FALSE
InitialUnmixSpanGates(template=IntermediateMetadata, gs=TheGS, subset="root", minpercentile=0.75, maxpercentile = 0.99)
```

```{r}
#| eval: FALSE
plot(TheGS)
```

```{r}
#| eval: FALSE
CheckUnmixingSpanGates(template=IntermediateMetadata, gs=TheGS, AdjustAll=TRUE)
```

```{r}
#| eval: FALSE
SaveName <- paste0("Aurora", Instrument, "_Day", Day, "UnmixingControls.gs") 
SaveHere <- file.path(OutputLocation, SaveName)
save_gs(TheGS, SaveHere)
```

```{r}
SaveName <- paste0("Aurora", Instrument, "_Day", Day, "UnmixingControls.gs") 
SaveHere <- file.path(OutputLocation, SaveName)
TheGS <- load_gs(SaveHere)
```

```{r}
plot(TheGS)
```

```{r}
RawValuesTemplate <- RetrieveAverageSignature(template=IntermediateMetadata, gs=TheGS, inverse.transform = TRUE)
RawValues <- BackgroundSubtraction(data=RawValuesTemplate, Normalize=FALSE, keepNegatives=FALSE)
```

# View Signatures

## Raw Values

```{r}
Plot <- VisualizeSignatures(columnname="Fluorophore", data=RawValues,
 Normalize=FALSE, characterColumns="Fluorophore")

plotly::ggplotly(Plot)
```

## Normalized Values

```{r}
Plot <- VisualizeSignatures(columnname="Fluorophore", data=RawValues,
 Normalize=TRUE, characterColumns="Fluorophore")

plotly::ggplotly(Plot)
```

# Unstained

```{r}
JustUnstaineds <- IntermediateMetadata |> dplyr::filter(Fluorophore %in% "Unstained")

Unstained <- LuciernagaIntegration(template=JustUnstaineds[1,],
 gs=TheGS, GuessSimilar=FALSE, Unstained=TRUE,
 inverse.transform=TRUE)[[1]]
```

```{r}
Plot <- Unstained$LuciernagaQC_Signatures + ggplot2::theme(legend.position = "none")
SaveName <- paste0("Aurora", Instrument, "_Day", Day, "AllAF.png") 
SaveAsThis <- file.path(OutputLocation, SaveName)
ggsave(SaveAsThis, units ="in", width = 12, height = 8)
```

```{r}
Plot <- Unstained$LuciernagaQC_AmalgamatedPlot + ggplot2::theme(legend.position = "none")
SaveName <- paste0("Aurora", Instrument, "_Day", Day, "AmalgamatedAF.png") 
SaveAsThis <- file.path(OutputLocation, SaveName)
ggsave(SaveAsThis, units ="in", width = 12, height = 8)
```


# Save the Dataset

```{r}
SaveName <- paste0("Aurora", Instrument, "_Day", Day, "SignatureValues.csv") 
SaveHere <- file.path(OutputLocation, SaveName)

write.csv(RawValues, SaveHere, row.names=FALSE)
```

```{r}
SaveName <- paste0("Aurora", Instrument, "_Day", Day, "Metadata.csv") 
SaveHere <- file.path(OutputLocation, SaveName)

write.csv(IntermediateMetadata, SaveHere, row.names=FALSE)
```

